Placing the spotted T Tauri star LkCa 4 on an HR diagram

### 2016-08-03 document/ms.tex
Settle on using f_Omega for the fill factor, add some text about observational fill factor

\providecommand{\eprint}[1]{\href{http://arxiv.org/abs/#1}{#1}}
\providecommand{\adsurl}[1]{\href{#1}{ADS}}
\newcommand{\name}{LkCa4 }
\newcommand{\project}[1]{\textsl{#1}}
%\def\vsini{$v\sin{i_*}$}

@@

\begin{abstract}
We interpret the high resolution near-IR spectra of \name.  We used the instrument IGRINS.
\end{abstract}

@@

Past spectroscopic detections of spots on young stars have focused on TW Hya, an active accretor, and DQ Tau, a close binary in which both components are accreting \citep{debes13,bary14}.  The WTTS \name \citep{herbig86,strom89a,downes88,strom89b} is an ideal exemplar for a spotted pre-MS star because it does not have any veiling \citep{hartigan95} or mid-IR or mm excess \citep{skrutskie90,andrews05,mccabe06,furlan06,luhman10,furlan11,harris12,marichalar12,howard13,buckle15}, is not actively accreting \citep{hartigan95,kenyon98,edwards06,cauley12,yang12,ardila13,castro13,loyd14,martinez15}, has no evidence for a close companion \citep{karr10,kraus11,daemgen15}\footnote{The status for wide companions is less clear; see \citet{stauffer91,itoh08,kraus09,kraus11}}, demonstrates large amplitude sinusoidal photometric variability indicative of large spots \citep{grankin08,xiao12}, has recently been examined with ZDI \citep{donati14}, and has large ($\Delta \teff \sim$500 K) disagreement about its effective temperature \citep{herczeg14}.  The spectrum of \name should be devoid of complicating factors like near-IR excess veiling and accretion excess.

@@

\subsection{IGRINS Spectroscopy}\label{sec:igrins} 
We acquired observations with the Immersion Grating Infrared Spectrograph, IGRINS \citep{gully12,park14} on the Harlan J. Smith Telescope at McDonald Observatory on 2015-11-18 $09^h$ UTC.  IGRINS is a high resolution near-infrared echelle spectrograph providing simultaneous $R\simeq45,000$ spectra over 1.48-2.48 \um.  The spectrograph has two arms with 28 orders in $H-$band and 25 orders in $K-$band.  The data were reduced with the Pipeline Package\footnote{\url{https://github.com/igrins/plp}}.  Telluric correction was performed by dividing the spectrum of \name by a spectrum of HR 1237, an A0V star observed immediately before \name, both at airmass 1.1.  The broad hydrogen lines in the A0V star produced broad flux excesses in the spectrum of \name.  No effort was made to remove these flux excesses, which affect several spectral orders.

@@

The Palomar/DBSP data was previously described in \citet{herczeg14}.

@@

An archive of 1173 $B$, $V$ and/or $R$ photometric monitoring visits was assembled from published and unpublished data sources spanning 31 years in 22 observing seasons.  Published data targeting \name\ is composed of 29 $BVR$ visits from 1985$-$1986 \citep{vrba93}, 26 $BVR$ epochs from 1990$-$1991 \citep{bouvier93}, 284 visits in $V$-band (278 with $B$ and 268 with $R$) from 1992 August - 2004 October \citep{grankin08}, and 10 $BVR$ visits from 2013 \citep{donati14}.  ASAS3 \citep{pojmanski04} acquired 63 $V-$band measurements from 2002$-$2004.  Integral-OMC obtained 138 $V-$band measurements from 2006$-$2008 \citep{garzon12}.  The AAVSO archive \citep{kafka16} includes 385 $V$, 23 $B$ and 10 $R$ measurements from 2013$-$2016.  Finally, unpublished data from the ASAS-SN survey \citep{shappee14}, were obtained from 2012 January -- 2016 March in 186 visits.  Figure \ref{fig:PhotTime} displays all the $V-$band photometry over the interval 1985-2016.

@@

\begin{eqnarray} \label{eqn:mix_M}
f_{\mathrm{mix}} = \Omega_a f(\teffa)  + \Omega_b f(\teffb)
\end{eqnarray}

The observed filling factor of starspots $c$ is simply the ratio:

\begin{eqnarray} \label{eqn:fill_factor}
c = \frac{\Omega_b}{\Omega_a + \Omega_b}
\end{eqnarray}

The starspot model includes two more free parameters than the standard \iancze\ model, namely $\teffb$ and $c$.  The addition of these two parameters makes the MCMC sampling much more correlated than it was before because the relative contribution of the starspot and ambient photosphere are nearly degenerate over small changes in effective temperature.  So the second important change to the spectral inference framework involves technical aspects of switching from sampling the nuisance and stellar parameters separately in a blocked Gibbs framework to ensemble sampling using \texttt{emcee} \citep{foreman13}.  The affine-invariant \texttt{emcee} ensemble sampler is more resilient to correlations among stellar temperatures and fill factor than the Metropolis-Hastings sampler used in Gibbs sampling.  The practical effect of this switch is that all 14 stellar and nuisance parameters are fit simultaneously in a single spectral order.  So stellar parameter estimates $\vT_{o}$ are unique for each spectral order $o$, whereas the \iancze\ strategy had the power to provide a single set of stellar parameters $\vT$ that was based on all $N_{\rm ord}$ spectral orders.  The $N_{\rm ord}$ sets of inferences on $\vT_{o}$ are combined by weighted averaging point estimates of results from reliable orders, which offers resilience to flagrant calibration artifacts or conspicuous model mismatches.  This process can be thought of as a coarse approximation to the much more sophisticated spectral line outlier rejection described in \iancze.

@@

We computed the median value and 5$^{th}$ and $95^{th}$ percentiles of burned-in subsets of the MCMC samples described above.  Overall these 26 sets of point estimates for $\vT_{m}$, $\vP_m$ show relatively good agreement, with exceptions.  We have little constraint on $\logg$ and $\Z$ since full-spectrum fitting tends to wash out the weak signal contained in line wings for spectra with a finite level of random noise and systematic error.  Meanwhile, strong metal lines can bias estimates of $\Z$ at the expense of estimates for $\teff$.  We get consistent estimates for $v_z \sim 15.6 \pm 0.6$ km/s and $\vsini \sim 28.0 \pm 0.9$ km/s, with the exception of two spectral orders at $6920 <\lambda \;(\mathrm{nm})< 7057$ and $8473 < \lambda \;(\mathrm{nm}) < 8706$.  These two outlying spectral orders converged to unreasonably large values for $\vsini$ and $a_{\rm G}$, the amplitude of the Gaussian process covariance kernel.  One of these orders has an unusually small value for effective temperature, $\teff\sim3500$, comparable to the lower limit of the employed model grid.  Visual comparison of the observed spectrum for a range of fits for these orders shows that no single emulated spectrum could reasonably fit their spectrum because there are numerous large spectral outliers.  The more-well-behaved spectral orders show effective temperature point estimates in the range $\teff=4000\pm130$ K.  Figure \ref{fig:SingleTeffvsOrder} displays the $\teff$ point estimates with 5$^{th}$ and $95^{th}$ percentile error bars placed at the central wavelength of each of the 26 spectral orders.

@@

We performed full-spectrum fitting on 32 of the 54 available IGRINS spectral orders, again fitting unique stellar parameters $\vt = (\teff, \logg, \Z)$ for each spectral order $m$.  We used the same analysis procedure as described for the ESPaDoNs spectra, with one exception.  For the IGRINS $K-$band, we employed an expanded search range for the effective temperature, $\teff \in [3000, 4200]$ K, since the IGRINS $H-$ band demonstrated some saturation at the $\teff=3500$ K lower bound.  No effort was made to shift the wavelength scales of the IGRINS data and \PHOENIX\ model into the same system, so the absolute values of $v_z$ are arbitrary shifted.

@@

Marginalized point estimates were computed for all 14 stellar and nuisance parameters by selecting the final 200 $\times$ 40 walkers = 8000 samples, and computing the median, $5^{th}$ and $95^{th}$ percentiles of the marginal distributions.  The fit quality was assessed by examining the consistency of the point estimates across the spectral orders.  The distribution of $v_z$ and $\vsini$ were informative for finding extremely poor fits.  The range for all 43 orders was $v_z = 12.4 \pm 2.6$ and $\vsini = 28.8 \pm 2.0$.  Spectral orders $m=91$ and $94$ had exceptionally large uncertainties on their values for $v_z$ and $\vsini$, which makes sense since these two orders are badly affected by telluric absorption artifacts.

@@

\begin{figure*}
	\centering
	\includegraphics[width=0.95\textwidth]{figures/violin_Teff_fill_order} 
	\caption{Violin plot of marginalized samples for $\teffa$, $\teffb$, and fill factor $c$, derived independently from full-spectrum fitting to 41 IGRINS orders.  Spectral orders show differing levels of constraint on the starspot and ambient photosphere properties, with some ($m=111$) consistent with negligible starspot emission, and others ($m=104, 102, 100$) showing tight constraints on the filling factor of starspots.  The starspot temperature is consistent with values even lower than 2700 K, the lower limit of the temperature range used.  $K-$band orders show lower estimates for the ambient photosphere, though some of these estimates are unreliable due to spectral line outliers}
	\label{fig:TwoTempResults}
\end{figure*}

@@

The reason for standardizing the flux is a matter of practicality: the number of PCA eigenspectra components in the spectral emulator scales steeply with the pixel-to-pixel variance of $f_{\lambda}(\{\vt_{\ast}\}^\textrm{grid})$, increasing the computational cost of spectral emulation.  The choice to standardize fluxes makes no difference for modeling a single photospheric component.  But for two-component photosphere models, the relative flux of the two model spectra needs to be accounted for to get an accurate estimate of the areal coverage fraction of the cool spots, $c \equiv \Omega_b/(\Omega_a+\Omega_b)$.  So we scale the mixture model in the following way:

@@

We are then tasked with computing an estimator, $\hat f(\vt_{\ast})$, for estimating the scale factor $q$ in-between model gridpoints.  The Right Thing to Do would be to follow \iancze by training a Gaussian process on the $f(\{\vt_{\ast}\}^\textrm{grid})$.  What We Actually Did was linearly interpolate between the model grid-points.  Interpolation can cause pile-up near model grid-points, as noted in \citet{cottaar14}, which motivated the spectral emulation procedure in \iancze.  We assume the interpolation of $\hat f$ is smooth enough that we will not see such pileups and if even we did see pileups, they would mostly be discernable in the distribution of samples in the starspot areal coverage fraction $c$.  One drawback of our estimator method compared to the Gaussian Process regression method is that we do not propagate the uncertainty associated with the absolute flux ratio interpolation into our estimate of $c$.  We assume this uncertainty is relatively small and can be ignored.

@@

Table \ref{tbl_history} lists measurements of LkCa4 from previous studies.
\input{tables/tbl_history.tex}

>>>>

\providecommand{\eprint}[1]{\href{http://arxiv.org/abs/#1}{#1}}
\providecommand{\adsurl}[1]{\href{#1}{ADS}}
\newcommand{\name}{LkCa 4 }
\newcommand{\project}[1]{\textsl{#1}}
%\def\vsini{$v\sin{i_*}$}

@@

\begin{abstract}
We interpret the high resolution near-IR spectra of LkCa 4.  We used the instrument IGRINS.
\end{abstract}

@@

Past spectroscopic detections of spots on young stars have focused on TW Hya, an active accretor, and DQ Tau, a close binary in which both components are accreting \citep{debes13,bary14}.  The WTTS \name \citep{herbig86,strom89a,downes88,strom89b} is an ideal exemplar for a spotted pre-MS star because it does not have any veiling \citep{hartigan95} or mid-IR or mm excess \citep{skrutskie90,andrews05,mccabe06,furlan06,luhman10,furlan11,harris12,marichalar12,howard13,buckle15}, is not actively accreting \citep{hartigan95,kenyon98,edwards06,cauley12,yang12,ardila13,castro13,loyd14,martinez15}, has no evidence for a close companion \citep{karr10,kraus11,daemgen15}\footnote{The status for wide companions is less clear; see \citet{stauffer91,itoh08,kraus09,kraus11,herczeg14}}, demonstrates large amplitude sinusoidal photometric variability indicative of large spots \citep{grankin08,xiao12}, has recently been examined with ZDI \citep{donati14}, and has large ($\Delta \teff \sim$500 K) disagreement about its effective temperature \citep{herczeg14}.  The spectrum of \name should be devoid of complicating factors like near-IR excess veiling and accretion excess.

@@

\subsection{IGRINS Spectroscopy}\label{sec:igrins} 
We acquired observations with the Immersion Grating Infrared Spectrograph, IGRINS \citep{gully12,park14} on the Harlan J. Smith Telescope at McDonald Observatory on 2015-11-18 $09^h$ UTC.  IGRINS is a high resolution near-infrared echelle spectrograph providing simultaneous $R\simeq45,000$ spectra over 1.48-2.48 \um.  The spectrograph has two arms with 28 orders in $H-$band and 25 orders in $K-$band.  The data were reduced with the Pipeline Package\footnote{\url{https://github.com/igrins/plp}}.  Telluric correction was performed by dividing the spectrum of \name by a spectrum of HR 1237, an A0V star observed immediately before \name, both at airmass 1.1.  The broad hydrogen lines in the A0V star produced broad flux excesses in the spectrum of LkCa 4.  No effort was made to remove these flux excesses, which affect several spectral orders.

@@

The Palomar/DBSP data was previously described in \citet{herczeg14}.  Briefly, these flux-calibrated spectra cover 3000-8700 $\AA$ at a spectral resolution $R\sim500-700$.

@@

An archive of 1173 $B$, $V$ and/or $R$ photometric monitoring visits was assembled from published and unpublished data sources spanning 31 years in 22 observing seasons.  Published data targeting \name is composed of 29 $BVR$ visits from 1985$-$1986 \citep{vrba93}, 26 $BVR$ epochs from 1990$-$1991 \citep{bouvier93}, 284 visits in $V$-band (278 with $B$ and 268 with $R$) from 1992 August - 2004 October \citep{grankin08}, and 10 $BVR$ visits from 2013 \citep{donati14}.  ASAS3 \citep{pojmanski04} acquired 63 $V-$band measurements from 2002$-$2004.  Integral-OMC obtained 138 $V-$band measurements from 2006$-$2008 \citep{garzon12}.  The AAVSO archive \citep{kafka16} includes 385 $V$, 23 $B$ and 10 $R$ measurements from 2013$-$2016.  Finally, unpublished data from the ASAS-SN survey \citep{shappee14}, were obtained from 2012 January -- 2016 March in 186 visits.  Figure \ref{fig:PhotTime} displays all the $V-$band photometry over the interval 1985-2016.

@@

\begin{eqnarray} \label{eqn:mix_M}
S_{\mathrm{mix}} = \Omega_a B(\teffa)  + \Omega_b B(\teffb)
\end{eqnarray}

where $B$ is the spectral radiance from emulated model spectra.  The observed filling factor of starspots $f_{\Omega}$ is simply the ratio:

\begin{eqnarray} \label{eqn:fill_factor}
f_{\Omega} = \frac{\Omega_b}{\Omega_a + \Omega_b}
\end{eqnarray}

Note that $f_{\Omega}$ represents an instantaneous, observational fill factor seen on one hemisphere, not $f_{\rm spot}$, the ``ratio of the spotted surface to the total surface areas'' \citep{somers15}.  In the limit of homogenously distributed spots, $f_{\Omega}$ tends to $f_{\rm spot}$, but in general $f_{\rm spot}$ is not a direct observable since some circumpolar regions of inclined stars will always face away from Earth; $f_{\Omega}$ can vary cyclicly though rotational modulation, whereas $f_{\rm spot}$ changes secularly through starspot evolution that is not yet understood.

The starspot model includes two more free parameters than the standard \iancze\ model, namely $\teffb$ and $f_{\Omega}$.  The addition of these two parameters makes the MCMC sampling much more correlated than it was before because the relative contribution of the starspot and ambient photosphere are nearly degenerate over small changes in effective temperature.  So the second important change to the spectral inference framework involves technical aspects of switching from sampling the nuisance and stellar parameters separately in a blocked Gibbs framework to ensemble sampling using \texttt{emcee} \citep{foreman13}.  The affine-invariant \texttt{emcee} ensemble sampler is more resilient to correlations among stellar temperatures and fill factor than the Metropolis-Hastings sampler used in Gibbs sampling.  The practical effect of this switch is that all 14 stellar and nuisance parameters are fit simultaneously in a single spectral order.  So stellar parameter estimates $\vt_{o}$ are unique for each spectral order $o$, whereas the \iancze\ strategy had the power to provide a single set of stellar parameters $\vt$ that was based on all $N_{\rm ord}$ spectral orders.  The $N_{\rm ord}$ sets of inferences on $\vt_{o}$ are combined by weighted averaging point estimates of results from reliable orders, which offers resilience to flagrant calibration artifacts or conspicuous model mismatches.  This process can be thought of as a coarse approximation to the much more sophisticated spectral line outlier rejection described in \iancze.

@@

We computed the median value and 5$^{th}$ and $95^{th}$ percentiles of burned-in subsets of the MCMC samples described above.  Overall these 26 sets of point estimates for $\vT_{o}$, $\vP_o$ show relatively good agreement, with exceptions.  We have little constraint on $\logg$ and $\Z$ since full-spectrum fitting tends to wash out the weak signal contained in line wings for spectra with a finite level of random noise and systematic error.  Meanwhile, strong metal lines can bias estimates of $\Z$ at the expense of estimates for $\teff$.  We get consistent estimates for $v_z \sim 15.6 \pm 0.6$ km/s and $\vsini \sim 28.0 \pm 0.9$ km/s, with the exception of two spectral orders at $6920 <\lambda \;(\mathrm{nm})< 7057$ and $8473 < \lambda \;(\mathrm{nm}) < 8706$.  These two outlying spectral orders converged to unreasonably large values for $\vsini$ and $a_{\rm G}$, the amplitude of the Gaussian process covariance kernel.  One of these orders has an unusually small value for effective temperature, $\teff\sim3500$, comparable to the lower limit of the employed model grid.  Visual comparison of the observed spectrum for a range of fits for these orders shows that no single emulated spectrum could reasonably fit their spectrum because there are numerous large spectral outliers.  The more-well-behaved spectral orders show effective temperature point estimates in the range $\teff=4000\pm130$ K.  Figure \ref{fig:SingleTeffvsOrder} displays the $\teff$ point estimates with 5$^{th}$ and $95^{th}$ percentile error bars placed at the central wavelength of each of the 26 spectral orders.

@@

We performed full-spectrum fitting on 32 of the 54 available IGRINS spectral orders, again fitting unique stellar parameters $\vt = (\teff, \logg, \Z)$ for each spectral order $o$.  We used the same analysis procedure as described for the ESPaDoNs spectra, with one exception.  For the IGRINS $K-$band, we employed an expanded search range for the effective temperature, $\teff \in [3000, 4200]$ K, since the IGRINS $H-$ band demonstrated some saturation at the $\teff=3500$ K lower bound.  No effort was made to shift the wavelength scales of the IGRINS data and \PHOENIX\ model into the same system, so the absolute values of $v_z$ are arbitrary shifted.

@@

Marginalized point estimates were computed for all 14 stellar and nuisance parameters by selecting the final 200 $\times$ 40 walkers = 8000 samples, and computing the median, $5^{th}$ and $95^{th}$ percentiles of the marginal distributions.  The fit quality was assessed by examining the consistency of the point estimates across the spectral orders.  The distribution of $v_z$ and $\vsini$ were informative for finding extremely poor fits.  The range for all 43 orders was $v_z = 12.4 \pm 2.6$ and $\vsini = 28.8 \pm 2.0$.  Spectral orders $o=91$ and $94$ had exceptionally large uncertainties on their values for $v_z$ and $\vsini$, which makes sense since these two orders are badly affected by telluric absorption artifacts.

@@

\begin{figure*}
	\centering
	\includegraphics[width=0.95\textwidth]{figures/violin_Teff_fill_order} 
	\caption{Violin plot of marginalized samples for $\teffa$, $\teffb$, and fill factor $f_{\Omega}$, derived independently from full-spectrum fitting to 41 IGRINS orders.  Spectral orders show differing levels of constraint on the starspot and ambient photosphere properties, with some ($o=111$) consistent with negligible starspot emission, and others ($o=104, 102, 100$) showing tight constraints on the filling factor of starspots.  The starspot temperature is consistent with values even lower than 2700 K, the lower limit of the temperature range used.  $K-$band orders show lower estimates for the ambient photosphere, though some of these estimates are unreliable due to spectral line outliers}
	\label{fig:TwoTempResults}
\end{figure*}

@@

The reason for standardizing the flux is a matter of practicality: the number of PCA eigenspectra components in the spectral emulator scales steeply with the pixel-to-pixel variance of $f_{\lambda}(\{\vt_{\ast}\}^\textrm{grid})$, increasing the computational cost of spectral emulation.  The choice to standardize fluxes makes no difference for modeling a single photospheric component.  But for two-component photosphere models, the relative flux of the two model spectra needs to be accounted for to get an accurate estimate of the areal coverage fraction of the cool spots, $f_{\Omega} \equiv \Omega_b/(\Omega_a+\Omega_b)$.  So we scale the mixture model in the following way:

@@

We are then tasked with computing an estimator, $\hat f(\vt_{\ast})$, for estimating the scale factor $q$ in-between model gridpoints.  The Right Thing to Do would be to follow \iancze by training a Gaussian process on the $f(\{\vt_{\ast}\}^\textrm{grid})$.  What We Actually Did was linearly interpolate between the model grid-points.  Interpolation can cause pile-up near model grid-points, as noted in \citet{cottaar14}, which motivated the spectral emulation procedure in \iancze.  We assume the interpolation of $\hat f$ is smooth enough that we will not see such pileups and if even we did see pileups, they would mostly be discernable in the distribution of samples in the starspot areal coverage fraction $f_{\Omega}$.  One drawback of our estimator method compared to the Gaussian Process regression method is that we do not propagate the uncertainty associated with the absolute flux ratio interpolation into our estimate of $f_{\Omega}$.  We assume this uncertainty is relatively small and can be ignored.

@@

Table \ref{tbl_history} lists measurements of \name from previous studies.
\input{tables/tbl_history.tex}
